Data Specialization for a Staged Lambda Calculus

### 2015-02-26 paper/splitting.tex
Attempting to add more "ease in" to the world 2 splitting correctness definition in 4.0.

%The output of splitting is simpler because it is {\em syntactically separated};
%The output of splitting is equivalent (to its input) because it can be
%evaluated to the same residual that would be produced by the semantics of
%\ref{sec:semantics}.

%More precisely

As described in \ref{sec:example},

For any $\coltwo e A$ which reduces to a residual $q$ (via $\diatwo e q$),
$e$ splits into a syntactically separated program $\mathcal{P}$ 
(via the new relation $e \splittwosym \mathcal{P}$) which is
reduced to a residual $q'$ (via the new relation $\mathcal{P} \sepredtwosym q'$), which is the same as $q$ up to an equivalence.
By ``syntactically separated'' we mean that $\mathcal{P}$ 
has the form $\pipeS p l r$, where  
$p$ (called the {\em precomputation}) is a monostage term representing all the
first-stage computations in $e$, 
and $l.r$ (called the {\em resumer}) is a monostage term representing all the
second-stage computations in $e$.  
$\sepredtwosym$ is defined by a single rule, which evaluates the precomputation and substitutes the resulting value in for $l$ in the resumer:
\[
\infer{\sepredtwo {\pipeS p l r} {[b/l]r}} {\reduce p b}
\]
We call $b$ the {\em boundary value} resulting from the split since it represents the data communicated between the first and second stage computations.
A summary of these operations and the relationship between them is provided on the right side of \ref{fig:splittingSummary}.

%Since terms at world \bbtwo\ can depend on terms at \bbonem\ (via
%\texttt{prev}), splitting is also defined on \bbonem\ terms.  However,
Defining the correctness requirements for splitting \bbonem\ terms is
challenging because unlike \bbtwo\ terms (which reduce to monostage
residuals that can be directly compared to the monostage result of
$\sepredtwosym$), \bbonem\ terms reduce to partial values and residual
tables that represent computations at both stages.
%One complication arises from the fact that \lang\ terms at world \bbonem\ produce multistage output.
For example, the \bbonem\ term:
\begin{lstlisting}
(next {1+2}, grnd{3+4})
\end{lstlisting}
reduces (via $\redonesym$) to the residual table and partial value:
\begin{lstlisting}
[yhat |-> 1+2] (next {yhat}, grnd{7})
\end{lstlisting}

To resolve the mismatch between the multi-stage computations
represented by residual tables and partial values and the desired monostaged
outputs of splitting,
% Since this output has interleaved stages, theres no way a syntactically separated term could directly reduce to it.
we choose to define a translation, called {\em masking}, written $\rtab \xi v \vsplito \mval i q$,
which converts residual tables and partial values into a syntactically separated {\em masked value}.
%version $\mathcal{V}$.
%As before, we enforce the ``syntactically separated'' property by defining $\mathcal{V}$ to have the form $\mval i q$ (called a {\em masked value})
The first component of the masked value, $i$, is a value holding the
result of all first stage computations in $v$.  The second component,
$q$, is a residual representing all second stage computations in $\rtab
\xi v$.

%where 
%$i$ is a monostage value representing the first stage part of $v$,
%and $q$ is a monostage term representing all second-stage computations in $\xi$ and $v$.

%Note that masking is a technical device used to state the correctness of splitting at \bbonem; 
%it is not part of the splitting algorithm itself.

@@

%Given the definition of masking, \bbonem\ terms split via the new
%relation $e \splitonesym \mathcal{P}$, where $\mathcal{P}$ must have
%the form $\pipeM c l r$
Similar to \bbtwo\ term splitting, $\mathcal{P}$ takes the form
$\pipeM c l r$ where $c$ (called the {\em combined term}) is a
monostage term encoding all the first-stage computations in $e$, and
$l.r$ (again called the {\em resumer}) is a monostage term encoding
all the second-stage computations in $e$.  $\sepredonesym$ is defined
by a single rule that yields masked values:
\[
\infer{\sepredone {\pipeM c l r} {\mval i {[b/l]r}}} {\reduce c {(i,b)}}
\]

>>>>

\textbf{Correctness condition for \bbtwo\ term splitting.}
Intuitively, as suggested in \ref{sec:example}, splitting a
\lang\ program (a \bbtwo\ term) should yield two monostage terms where
evaluating the second term (corresponding to second stage
computations) on the output of the first (first stage computations),
yields the same result as direct evaluation of the \lang\ program.
More precisely, for any $\coltwo e A$ which reduces to a residual $q$
(via $\diatwo e q$), $e$ splits into a syntactically separated program
$\mathcal{P}$ (via the new relation $e \splittwosym \mathcal{P}$)
which is reduced to a residual $q'$ (via the new relation $\mathcal{P}
\sepredtwosym q'$), which is the same as $q$ up to an equivalence.  By
``syntactically separated'' we mean that $\mathcal{P}$ has the form
$\pipeS p l r$, where $p$ (called the {\em precomputation}) is a
monostage term representing all the first-stage computations in $e$,
and $l.r$ (called the {\em resumer}) is a monostage term representing
all the second-stage computations in $e$.  $\sepredtwosym$ is defined
by a single rule, which evaluates the precomputation and substitutes
the resulting value in for $l$ in the resumer:
\[
\infer{\sepredtwo {\pipeS p l r} {[b/l]r}} {\reduce p b}
\]
We call $b$ the {\em boundary value} resulting from the split since it represents the data communicated between the first and second stage computations.
A summary of these operations and the relationship between them is provided on the right side of \ref{fig:splittingSummary}.

\textbf{Correctness condition for \bbonem\ term splitting.} Defining
the correctness requirements for splitting \bbonem\ terms is
challenging because unlike \bbtwo\ terms (which reduce to monostage
residuals that can be directly compared to the monostage result of
$\sepredtwosym$), \bbonem\ terms reduce to partial values and residual
tables that represent computations at both stages.
%One complication arises from the fact that \lang\ terms at world \bbonem\ produce multistage output.
%For example, the \bbonem\ term:
%\begin{lstlisting}
%(next {1+2}, grnd{3+4})
%\end{lstlisting}
%reduces (via $\redonesym$) to the residual table and partial value:
%\begin{lstlisting}
%[yhat |-> 1+2] (next {yhat}, grnd{7})
%\end{lstlisting}

To resolve the mismatch between the multi-stage computations
represented by residual tables and partial values and the desired
monostaged outputs of splitting, we choose to define a translation,
called {\em masking}, written $\rtab \xi v \vsplito \mval i q$, which
converts residual tables and partial values into a syntactically
separated {\em masked value}.  The first component of the masked
value, $i$, is a value holding the result of all first stage
computations in $v$.  The second component, $q$, is a residual
representing all second stage computations in $\rtab \xi v$.

@@

Similar to \bbtwo\ term splitting, $\mathcal{P}$ takes the form
$\pipeM c l r$ where $c$ (called the {\em combined term}) is a
monostage term encoding all the first-stage computations in $e$, and
$l.r$ (again called the {\em resumer}) is a monostage term encoding
all the second-stage computations in $e$.  $\sepredonesym$ is defined
by a single rule that yields masked values:
\[
\infer{\sepredone {\pipeM c l r} {\mval i {[b/l]r}}} {\reduce c {(i,b)}}
\]
